Quantifying the Importance of Change for Understanding Differences in Personal Culture

### 2024-03-14 manuscript/manuscript.tex
kkedits314

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fontspec} \setmainfont{Times New Roman} \setlength{\parindent}{3em} \usepackage{caption} \usepackage{setspace}\doublespacing \usepackage{endnotes} \renewcommand{\enotesize}{\normalsize} \let\footnote=\endnote
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quantifying The Importance of Change for Understanding Differences in Personal Culture},
  pdfauthor={Achim Edelmann; Kevin Kiley; Turgut Keskintürk; Isabella Bouklas; Stephen Vaisey},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Quantifying The Importance of Change for Understanding
Differences in Personal Culture}
\author{Achim Edelmann\footnote{médialab, Science Po} \and Kevin
Kiley\footnote{Department of Sociology and Anthropology, North Carolina
  State University} \and Turgut Keskintürk\footnote{Department of
  Sociology, Duke University} \and Isabella
Bouklas\footnote{Department of Sociology, Duke University} \and Stephen
Vaisey\footnote{Department of Sociology, Duke University}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
When researchers want to understand differences in personal culture---a
person's attitudes, beliefs, values, and practices--how much attention
should they pay to adult experience? Recent work has reached
substantially different conclusions on this question. We argue that this
disagreement is an unintended consequence of the ``tournament of
models'' approach researchers have used, which focuses on whether people
change and not how much they change. To advance the theoretical debate,
we refocus attention on the relative importance of personal change over
time for explaining differences between people. We introduce a new
measure for quantifying the proportion of systematic variance in panel
data attributable to intrapersonal change. Applying this measure to 609
items from seven surveys in five countries, we find that although
intrapersonal change is common, it is generally small in magnitude. As
an extension of the theoretical debate, we demonstrate that this measure
provides new insights when comparing social groups, showing that
intrapersonal change is less common among U.S. college graduates than
among those without a college degree. Our findings provide a new
perspective on several important theoretical debates, as well as a tool
to address new questions.
\end{abstract}

@@

Does personal culture---a person's attitudes, beliefs, values, and
practices---change over the life course, or is it largely fixed by
adulthood? This question underlies an important contemporary debate in
sociology (Kiley and Vaisey 2020; Lersch 2023; Lizardo 2017) and has
deep roots in seemingly contradictory theoretical perspectives. For
example, pragmatist theories of action claim that changes in social
environments cause people to adapt their views and make new cultural
meanings (Gross 2009; Swidler 2001), while Bourdieusian practice
theories argue that the ``past conditions of production'' leave a mark
on people's personal culture that lasts throughout their lives (Bourdieu
1990). Models of social influence assume that people adapt their culture
in the face of new information (Goldberg and Stein 2018), while the
emphasis on cohort effects in models of aggregate social and cultural
change requires them to be open to change while young but become fairly
resistant to it as they age (Ryder 1965). Finally, life course theories
posit important changes over time as people advance through important
transitions in their lives (Bardi et al. 2009; Elder, Johnson, and
Crosnoe 2003).

Because all of these perspectives have some empirical support, the
theoretical debate is not about whether the processes they posit exist
at all, but about their relative contribution to explaining the cultural
differences we see in the world. Nevertheless, researchers have
struggled to reach a consensus on the importance of cultural change
during adulthood. Over the span of a few years, most apparent changes in
adults' survey responses appear to be transitory, with little evidence
of persistent change (Kiley and Vaisey 2020; Vaisey and Kiley 2021).
This suggests that change during adulthood may not be a major factor in
explaining contemporary cultural divides. By contrast, when we consider
a longer time horizon, there is evidence that adults make at least some
persistent changes (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach to adjudicate
different theoretical perspectives (Lersch 2023: p.~228). That is,
researchers have relied on model selection criteria---primarily the
Bayesian Information Criterion (Raftery 1995)---to sort variables into
the ``no change'' or ``any change'' piles. The size of these piles then
serves as the primary evidence for the truth of the theory (Kiley and
Vaisey 2020; Lersch 2023; Vaisey and Kiley 2021).

We argue, however, that asking \emph{whether} people change on an item
is an unintended source of confusion. Being able to detect \emph{any}
change on an item depends not only on sample populations, measurement,
statistical methods, and power, but also on definitions of what counts
as ``change.'' These differences can limit the possibility of agreement
on the assignment of a variable to the right pile. Moreover, asking
whether there is evidence of change reduces to, ``Is there \emph{any}
evidence of change in this item?'' But such binary questions are
ill-suited to address the core of the theoretical debate: determining
\emph{how much} the process of intrapersonal change contributes to
cultural differences.

Of course neither pre-adult socialization nor the contemporary social
context alone can explain the full range of cultural differences. The
relevant question is about the \emph{relative} contributions of these
processes. In the current debate, for a single construct (e.g., support
for gay rights), model comparison approaches can only help decide
\emph{whether} intrapersonal change is happening, not about how much.
This also prevents investigating under what conditions, across which
groups, and in what domains we might see differences in the amount of
intrapersonal change. These questions are about \emph{degree}, not
existence. Taking this debate forward therefore requires precise
measurement rather than declarations of victory for a particular
perspective.

In this paper, we move beyond the ``tournament of models'' approach and
introduce a new method for quantifying the relative contributions of
interpersonal differences and intrapersonal change for a single item. We
use 609 items from seven panel surveys from five countries to quantify
the proportion of variance that is attributable to systematic
intrapersonal change over the study period rather than to stable
interpersonal differences.

We observe similar results across all datasets despite their varied
social contexts and duration. In general, we find that intrapersonal
change accounts for a small amount of variance in personal culture
survey items. Of course, any question about whether a given amount of
intrapersonal change is ``meaningful'' or ``important'' is substantive
and theoretical. Nevertheless, on many measures (some observed for over
a decade), we do not see enough change to believe that intrapersonal
change processes play a substantial role in explaining the differences
we observe between adults. Notable, interesting, and important
exceptions exist, but they are departures from the overall pattern.

In addition to advancing a theoretical debate, our method can also help
address specific substantive questions. To illustrate how it might be
used by a researcher with a specific topical interest, we investigate
differences in the relative importance of intrapersonal change for
people with and without a college degree. We find a smaller amount of
systematic change among college-educated respondents, suggesting that
college crystallizes one's personal culture rather than fostering an
openness to new information. This offers a new empirical basis to
theorize about the role of education in influencing personal culture.

Taken together, our goal in this paper is to advance---and, hopefully,
transcend---debates about \emph{whether} adults change. People do
change, at least a little, on most things. Our method can help quantify
exactly \emph{how much}.

@@

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items did prefer the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they were more consistent with
zero durable change. Questions with more evidence of durable change
included salient issues like gay marriage and questions tapping
``public'' statements or behaviors such as partisan identification and
religious service attendance. There was also more evidence of durable
change among early adults (people ages 18-30) than among the rest of the
population. Overall, the researchers concluded that ``results ultimately
suggest that real, persistent attitude change is an uncommon phenomenon
among adults'\,' (Kiley and Vaisey 2020: p.~500; see also Vaisey and
Kiley 2021). This lack of durable change is consistent with other recent
findings that cohort replacement plays a somewhat larger role than
period effects in explaining differences in personal culture (Vaisey and
Lizardo 2016).

@@

Drawing on these findings, Lersch (2023) challenged the SDM and AUM as a
``needless dichotomy,'' proposing the ``Life Course Adaption Model''
(LCAM) as an alternative. This model draws on the life course
perspective to model personal culture as a different linear trajectory
over the duration of a panel for each respondent. In doing this, Lersch
rectified two shortcomings of the AUM and SDM. First, the AUM, as
described by Kiley and Vaisey, posits that changes follow a Markov
process where responses at time \(t\) are a function of responses at
time \(t-1\) but not earlier time points. However, earlier life
experiences can (directly and indirectly) mold personal culture when
transitioning to new social roles or into new environments, even if
their initial impact is delayed. For example, childhood events might
influence views on family structures later when individuals form their
own families. Therefore, the LCAM considers influences from earlier than
just \(t-1\) on responses at time \(t\).

Second, Kiley and Vaisey's analysis of the AUM and SDM is based on
three-wave panel data over four years, which might not be intensive or
extensive enough to adjudicate the two models. Lersch evaluates the LCAM
against the AUM and SDM using panel data spanning a wider duration (from
3 to 36 years) and more waves (3 to 18), offering a better chance to
observe durable change.

When Lersch compared the LCAM to the AUM and SDM on data from five
countries, 297 of the 428 questions he analyzed preferred the LCAM,
suggesting that we observe some linear change for adults on most items.
The SDM was preferred on 112 items, and the rest did not yield
conclusive results. No questions favored the AUM. He concludes that
``new experiences over the life course {[}\ldots{]} can persistently
move individuals' personal culture in novel directions'' (Lersch 2023:
p.~243-244).

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of durable change among adults, a pattern consistent with
Lersch's results. Also Lersch (2023) found evidence of durable change on
most items, but these changes were small. On average, people changed
only about .07 standard deviations over 10 years. This is consistent
with Kiley and Vaisey's finding that durable change is often small.
Moreover, a quarter of the items Lersch studied still preferred the SDM,
suggesting that even with different assumptions about change and more
extensive data, many questions still are consistent with a model that
allows zero intrapersonal change.

@@

To this point, the debate has been about whether people \emph{ever}
change. As we argued above, this is rarely the right question since the
answer is almost always ``yes.'' Over sufficient time and in a large
enough sample, researchers will likely observe some evidence for durable
individual change. A lack of evidence for change may be due to a poor
survey question, a low resolution in response options, or that the
question simply was not asked for long enough or to enough people.
Conversely, finding evidence for some change tells us only that it was
``not zero.'' From that alone, we learn little about how much
intrapersonal change has happened in a population.

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM does this by modeling each individual as
following a personal linear trajectory. But rather than debating whether
this model fits a particular data set better than a model that assumes
that people never change, it would be better to quantify these
individual changes and compare them to the stable differences that exist
between people. In other words, the focus should shift from asking, ``do
people change?'' to asking, ``what are the relative contributions of
change and pre-existing differences for explaining variance in personal
culture?''

We propose an approach---which we formalize below---that offers such a
quantification. Our measure relates two variance components that each
reflect a combination of relevant theoretical processes. The first
component, stable interpersonal differences, reflects the accumulated
experiences of people prior to entering the panel survey. Lersch (2023:
24) calls this ``early imprinting.'' While commonly associated with
experiences during a formative period that result in settled
dispositions, this variation could also reflect experiences that
happened at any time as long as they predate the panel and consistently
affect subsequent responses. For instance, for those people entering the
panel post-retirement, this ``imprinting'' might reflect this pivotal
life transition. Consequently, this component also reflects variation in
individuals' social roles or statuses at the start of the panel that
were important in shaping their dispositions.

The second component, the amount of intrapersonal linear change people
make during the panel, captures durable changes in personal culture over
time. This captures the set of processes collectively called
``persistent change'' or ``adaption'' by Lersch and ``active updating''
by Kiley and Vaisey. Lersch attributes these changes to social triggers
such as moving into a new environment or adopting new social roles
(although he does not measure these directly). They might also reflect
the diffusion of new cultural forms across social networks, cues from
political elites or otherwise culturally influential leaders, the
emergence of issues in politics or culture, or large-scale social
shifts.

@@

@@

\hypertarget{life-course-adaption-model}{%
\subsection{Life Course Adaption
Model}\label{life-course-adaption-model}}

The estimation procedure begins with Lersch's (2023) Life Course
Adaption Model, which formalizes survey responses at time \(t\) as a
function of individual-level random intercepts and slopes for survey
age. In some contexts, this model is called a mixed-effects growth curve
model. This model assumes a set of propositions about change that
reflect the theoretical debate to this point. First, consistent with the
settled disposition model, it assumes that people start the survey with
cultural differences, modeled as random intercepts for each respondent.
Second, it assumes that people change over time, taking the form of
random slopes for each respondent as a linear function of time. Third,
it assumes that people deviate around this baseline randomly over time,
reflecting ``fluctuation'' or short-term non-persistent change.
Formally, this can be written as

@@

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then use these models to
derive two measures of the components we outlined above: pre-existing
interpersonal differences and intrapersonal change over time.

The third component, which is the proportion of variance attributable to
measurement error or transient fluctuations, is what's left, such that
these three numbers sum to 1. As we discussed above, this component is
not of core interest here.

As a second step in our analysis, we use linear regression to conduct a
meta-analysis, modeling \(\omega\) as a function of features of the
question and panel. This allows us to explore how these features are
associated with the degree of intrapersonal change. These covariates
include the number of response options (measured as 2, 3-5, and more
than 5), the specific survey, the log of the number of participants, the
number of waves observed, the time period (i.e., decade) the question
was first asked, and the total duration of time the question was asked.

@@

Although we believe it is a major advance on previous work, the approach
we outline here has important limitations. We allocate all systematic
variance to one of two sets of theoretical processes: intrapersonal
change and interpersonal differences at baseline. Our approach does not
quantify the proportion of people who ``change,'' nor can we be sure
that the amount of interpersonal change we detect is driven by many
people making small changes or a few people making large changes.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fontspec} \setmainfont{Times New Roman} \setlength{\parindent}{3em} \usepackage{caption} \usepackage{setspace} \usepackage{endnotes} \renewcommand{\enotesize}{\normalsize} \let\footnote=\endnote
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quantifying the Importance of Change for Understanding Differences in Personal Culture},
  pdfauthor={Achim Edelmann; Kevin Kiley; Turgut Keskintürk; Isabella Bouklas; Stephen Vaisey},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Quantifying the Importance of Change for Understanding
Differences in Personal Culture}
\author{Achim Edelmann\footnote{médialab, Science Po} \and Kevin
Kiley\footnote{Department of Sociology and Anthropology, North Carolina
  State University} \and Turgut Keskintürk\footnote{Department of
  Sociology, Duke University} \and Isabella
Bouklas\footnote{Department of Sociology, Duke University} \and Stephen
Vaisey\footnote{Department of Sociology, Duke University}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Work in the sociology of culture has converged on a general conclusion
that while people sometimes make substantial shifts in their personal
culture over time, most elements of personal culture are characterized
by low rates of persistent change during adulthood. Recent developments
have begun to quantify change over time and initial differences, but it
is unclear how to use these components to understand the relative
importance of change, especially when change is modeled non-linearly. To
advance this debate, we introduce a measure for quantifying the
proportion of systematic variance in panel data attributable to
intrapersonal change. Applying this measure to 609 items from seven
surveys in five countries, we find that although intrapersonal change is
common, it does not explain a large share of systematic variance. As an
extension of the theoretical debate, we demonstrate that this measure
provides new insights when comparing social groups, showing that
intrapersonal change is less common among U.S. college graduates than
among those without a college degree. Our findings provide a new
perspective on several important theoretical debates, as well as a tool
to address new questions.
\end{abstract}

\doublespacing

@@

An important contemporary debate in sociology centers on the importance
of intrapersonal change for understanding differences in ``personal
culture'' -- attitudes, beliefs, values, and practices (Kiley and Vaisey
2020; Lersch 2023; Lizardo 2017; Vaisey and Lizardo 2016;
\textbf{underwood2021?}). This question has deep roots in seemingly
contradictory theoretical perspectives. For example, pragmatist theories
of action claim that changes in social environments cause people to
adapt their views and make new cultural meanings (Gross 2009; Swidler
2001), while Bourdieusian practice theories argue that the ``past
conditions of production'' leave a mark on people's personal culture
that lasts throughout their lives (Bourdieu 1990). Models of social
influence assume that people adapt their culture in the face of new
information (Goldberg and Stein 2018), while the emphasis on cohort
effects in models of aggregate social and cultural change requires them
to be open to change while young but become fairly resistant to it as
they age (Ryder 1965). Finally, life course theories posit important
changes over time as people advance through important transitions in
their lives, but that certain life stages have more lasting influence
than others (Bardi et al. 2009; Elder, Johnson, and Crosnoe 2003).

Recent work on this topic has converged on a general conclusion that
people do sometimes make substantial shifts in their personal culture
over time but that most elements of personal culture are characterized
by low rates of persistent change during adulthood (Kiley and Vaisey
2020; Lersch 2023). Until recently, this work has adopted a ``tournament
of models'' approach to adjudicating different theoretical perspectives
(Lersch 2023: p.~228). In this framework, researchers formalized
different models of change over time and relied on model selection
criteria---primarily the Bayesian Information Criterion (Raftery
1995)---to classify different questions as demonstrating either ``no
change'' or ``any change.'' The size of these piles then serves as the
primary evidence for the truth of different arguments (Kiley and Vaisey
2020; Vaisey and Kiley 2021).

However, as consensus in the overall picture emerges, researchers are
now turning toward understanding the conditions that produce persistent
change in personal culture, including age, education, and life-course
transitions; whether certain kinds of people are more likely to make
persistent changes in personal culture; and whether certain kinds of
personal culture are more likely to be characterized by persistent
change or stability. The tournament of models approach, useful for the
broad question of whether change happens, is ill-suited to addressing
these more theoretically grounded questions.

In a recent paper, Lersch (2023) takes an important step forward by
devising a Life Course Adaption Model (LCAM), a mixed-effects growth
curve model that quantifies various components of cultural difference:
early imprinting, persistent change, biographical experience, and
fluctuation. In quantifying these aspects of cultural difference, the
LCAM lets researchers address a broader range of questions than the
single question of whether a measure is characterized by stability or
change. However, because the LCAM produces so many different measures,
it is not obvious how to use these components to understand the relative
importance of change, espeically when change is measured in non-linear
ways.

In this paper, we build on the LCAM by outlining a single metric that
combines diverse components of the LCAM to quantify the relative
importance of intrapersonal change and interpersonal difference for
explaining variation in a population for a variable over time. Our
approach isolates the ``systematic'' components of variance (i.e., it
removes random change and measurement error), to provide a single
succinct summary of how much systematic variation can reasonably be
attributed to people making persistent changes over time. We demonstrate
the utility of this quantity in two ways.

First, we use 609 items from seven panel surveys from five countries to
quantify the proportion of variance that is attributable to systematic
intrapersonal change over the study period rather than to interpersonal
differences at baseline. This allows us to generate an overall picture
of how much of the total systematic variation in personal culture change
tends to explain. These findings largely echo previous findings but
enable clearer statements about the relative importance of change across
questions, rather than binary claims about whether change ``happens'' or
specific quantification of how much change occurs.

Second, to illustrate how this quantity might be used by a researcher
with a specific topical interest, we investigate differences in the
relative importance of intrapersonal change for people with and without
a college degree. We find that change explains a smaller proportion of
systematic variance among college-educated respondents, suggesting that
college crystallizes one's personal culture rather than fostering an
openness to new information. This offers a new empirical basis to
theorize about the role of education in influencing personal culture.

@@

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items did prefer the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they were more consistent with
zero durable change. Questions with more evidence of durable change
included salient issues like gay marriage and questions tapping
``public'' statements or behaviors such as partisan identification and
religious service attendance. There was also more evidence of durable
change among early adults (people ages 18-30) than among the rest of the
population. Overall, the researchers concluded that ``results ultimately
suggest that real, persistent attitude change is an uncommon phenomenon
among adults'\,' (Kiley and Vaisey 2020: p.~500; see also Vaisey and
Kiley 2021). This lack of durable change is consistent with other recent
findings that cohort replacement plays a larger role than period effects
in explaining differences in personal culture (Vaisey and Lizardo 2016).

@@

Drawing on these findings, Lersch (2023) challenged the SDM and AUM as a
``needless dichotomy,'' proposing the ``Life Course Adaption Model''
(LCAM) as an alternative. This model draws on the life course
perspective to model personal culture as a different linear
trajectory\footnote{In supplemental materials, Lersch models age using
  splines as well, with results generally showing the same conclusions.}
over the duration of a panel for each respondent. In doing this, Lersch
rectified three shortcomings of the AUM and SDM. First, the AUM, as
described by Kiley and Vaisey, posits that changes follow a Markov
process where responses at time \(t\) are a function of responses at
time \(t-1\) but not earlier time points. However, earlier life
experiences can (directly and indirectly) mold personal culture when
transitioning to new social roles or into new environments, even if
their initial impact is delayed. For example, childhood events might
influence views on family structures later when individuals form their
own families. Therefore, the LCAM considers influences from earlier than
just \(t-1\) on responses at time \(t\).

Second, Kiley and Vaisey's conclusions about the AUM and SDM were based
on examining three-wave panel data over four years, which might not be
intensive or extensive enough to adjudicate the two models. Lersch
evaluates the LCAM against the AUM and SDM using panel data spanning a
wider duration (from 3 to 36 years) and more waves (3 to 18), offering a
better chance to observe durable change.

Third, and most importantly, the LCAM quantifies various components of
cultural difference, allowing researchers to ask a more varied set of
questions than simply whether people change. For example, researchers
can use the LCAM framework to quantify the magnitude of change in a
measure of culture over time (i.e., how much do people change on
average), the variance in age-based slopes over time (i.e., do people
tend to change in similar ways as they age), or how much personal
culture tends to fluctuate around these age-based trends (how consistent
are responses over time).

When Lersch compared the LCAM to the AUM and SDM on data from five
countries, 198 of the 428 questions he analyzed preferred the LCAM and
113 preferred the AUM, suggesting that we observe some change for adults
on most items. The SDM was preferred on 93 items, and the rest did not
yield conclusive results. He concludes that ``new experiences over the
life course {[}\ldots{]} can persistently move individuals' personal
culture in novel directions'' (Lersch 2023: p.~243-244).

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of persisting change among adults, a pattern consistent
with Lersch's results. Also Lersch (2023) found evidence of persisting
change on most items, but these changes were small. On average, people
changed only about .07 standard deviations over 10 years. This is
consistent with Kiley and Vaisey's finding that persisting change is
often small. Moreover, a quarter of the items Lersch studied still
preferred the SDM, suggesting that even with different assumptions about
change and more extensive data, many questions still are consistent with
a model that allows zero intrapersonal change.

@@

To this point, the debate has broadly centered on drawing conclusions
about whether people \emph{ever} change. But this is rarely the right
question since the answer is almost always ``yes.'' Over sufficient time
and in a large enough sample, researchers will likely observe some
evidence for durable change in the population. A lack of evidence for
change may be due to a poor survey question, a low resolution in
response options, or that the question simply was not asked for long
enough or to enough people. Conversely, finding evidence for some change
tells us only that it was ``not zero.'' From that alone, we learn little
about how much intrapersonal change has happened in a population.

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM does this by modeling each individual as
following a personal linear trajectory. In doing so, the LCAM opens the
door to a variety of questions. We assert that among the many things
that can be done with the LCAM, it provides the opportunity to shift
from asking, ``do people change?'' to asking, ``what are the relative
contributions of change and pre-existing differences for explaining
variance in personal culture?''

Our approach---which we formalize below---offers such a quantification.
Our measure relates two variance components that each reflect a
combination of relevant theoretical processes. The first component,
stable interpersonal differences, reflects the accumulated experiences
of people prior to entering the panel survey. Lersch (2023: 24) calls
this ``early imprinting.'' While commonly associated with experiences
during a formative period that result in settled dispositions, this
variation could also reflect experiences that happened at any time as
long as they predate the panel and consistently affect subsequent
responses. For instance, for those people entering the panel
post-retirement, this ``imprinting'' might reflect this pivotal life
transition. Consequently, this component also reflects variation in
individuals' social roles or statuses at the start of the panel that
were important in shaping their dispositions.

The second component, the amount of intrapersonal change people make
during the panel, captures persisting changes in personal culture over
time. This component captures the set of processes collectively called
``persistent change'' or ``adaption'' by Lersch and ``active updating''
by Kiley and Vaisey. Lersch attributes these changes to social triggers
such as moving into a new environment or adopting new social roles
(although he does not measure these directly). They might also reflect
the diffusion of new cultural forms across social networks, cues from
political elites or otherwise culturally influential leaders, the
emergence of issues in politics or culture, or large-scale social
shifts.

@@

\hypertarget{formalization}{%
\subsection{Formalization}\label{formalization}}

The starting point for our quantity is Lersch's (2023) Life Course
Adaption Model, which formalizes survey responses at time \(t\) as a
function of individual-level random intercepts and slopes for survey
age. In some contexts, this model is called a mixed-effects growth curve
model. This model assumes a set of propositions about change that
reflect the theoretical debate to this point. First, consistent with the
settled disposition model, it assumes that people start the survey with
cultural differences, modeled as random intercepts for each respondent.
Second, it assumes that people change over time, taking the form of
random slopes for each respondent as a linear function of time. Third,
it assumes that people deviate around this baseline randomly over time,
reflecting ``fluctuation'' or short-term non-persistent change.
Formally, this can be written as

We then use these models to derive two measures of the components we
outlined above: pre-existing interpersonal differences and intrapersonal
change over time.

The third component, which is the proportion of variance attributable to
measurement error or transient fluctuations, is what's left, such that
these three numbers sum to 1. As we discussed above, this component is
not of core interest here, so we set it aside in our analysis.

\hypertarget{bootstrapping-for-confidence-intervals}{%
\subsection{Bootstrapping for Confidence
Intervals}\label{bootstrapping-for-confidence-intervals}}

{[}Explain how we do that here{]}

@@

@@

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then derive \(\omega\)
for each of these items.

As a second step in our analysis, we use linear regression to model
\(\omega\) as a function of features of the question and panel. This
allows us to explore how these features are associated with the degree
of intrapersonal change. These covariates include the number of response
options (measured as 2, 3-5, and more than 5), the specific survey, the
log of the number of participants, the number of waves observed, the
time period (i.e., decade) the question was first asked, and the total
duration of time the question was asked.

@@

Although we believe the approach we outline here advances previous work,
it has important limitations. We allocate all systematic variance to one
of two sets of theoretical processes: intrapersonal change and
interpersonal differences at baseline. Our approach does not quantify
the proportion of people who ``change,'' nor can we be sure that the
amount of interpersonal change we detect is driven by many people making
small changes or a few people making large changes.
